# Exploratory data analysis

## Description of the state of the spread

Provide a high level description of the state of the spread. In particular, include

  * the number of days that have passed since the first confirmed case/death,
  * the current stage for confirmed cases/deaths/mortality (i.e., ratio of deaths to confirmed cases).
  
You can:

  * use either words, or tables, or figures, or a bit of each,
  * group your analysis per country, or region, or a bit of each.
  
  
```{r, warning=FALSE, message=FALSE, echo = FALSE}
library(tidyverse)
# all the files from the data folder
files <- "data" %>%
  here::here() %>%
  list.files(full.names = TRUE)

# latest data file
latest_data <- files %>%
  str_subset("jh_covid19_data") %>%
  str_extract("\\d{4}-\\d{2}-\\d{2}") %>%
  max(na.rm = TRUE)

# load the latest data
covid19_data <- files %>%
  str_subset("jh_covid19_data") %>%
  str_subset(latest_data) %>%
  read_csv()

# load the worldbank data
worldbank_data <- files %>%
  str_subset("jh_add_wbank_data") %>%
  read_csv()

# join the covid19 and worldbank datasets
covid19_data <- covid19_data %>%
  inner_join(
    worldbank_data,
    by = "iso3c"
  )
 
```

```{r, warning=FALSE, message=FALSE, echo = FALSE}
##load the library
library(tidyverse)
library(kableExtra)

```


```{r fig-1, fig.cap="Figure 1", out.width="80%", fig.asp=.75,fig.align='center'}


## Analyze worlwide


covid19_data %>% group_by(date) %>% summarize(Confirmed= sum(confirmed), Deaths = sum(deaths)) %>% ggplot() +
geom_line(aes(x=date, y= Confirmed, col="Confirmed"))+
geom_line(aes(x=date, y=Deaths, col="Deaths")) +
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d" )+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Higly increase for the confirmed case", subtitle = "Worldwide representation", x= "Date",y= "People", caption = "data from jh_covid19_data_2020-04-06.csv")

````

```{r fig-2, fig.cap="Figure 2", out.width="80%", fig.asp=.75,fig.align='center'}

covid19_data <- covid19_data %>% mutate(ratio=(deaths/confirmed))
covid19_data$ratio[is.nan(covid19_data$ratio)]<-0

covid19_data %>% group_by(date) %>% summarize(Mean_Ratio = mean(ratio)) %>% ggplot() +
  geom_line(aes(x=date, y=Mean_Ratio))  +
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d" ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
   labs(title = "Highly increase in the ratio deaths/confirmed", subtitle = "Worldwide representation", x= "Date",y= "Ratio", caption = "data from jh_covid19_data_2020-04-06.csv")

```

```{r fig-3, fig.cap="Figure 3", out.width="80%", fig.asp=.75,fig.align='center'}

 covid19_data %>% group_by(date,region) %>% summarize(Confirmed= sum(confirmed)) %>% ggplot(aes(x=date, y= Confirmed, colour= region)) +
  geom_line() +
   labs(title = "Flat increase and early deperature in confirmed case in the East Asia & Pacific", subtitle = "Worldwide representation group by region", x= "Date",y= "Confirmed case", caption = "data from jh_covid19_data_2020-04-06.csv")+
   scale_x_date(date_breaks = "1 week", date_labels = "%b %d" ) +
   theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r fig-4, fig.cap="Figure 4", out.width="80%", fig.asp=.75,fig.align='center'}

 covid19_data %>% group_by(date,region) %>% summarize(Deaths= sum(deaths)) %>% ggplot(aes(x=date, y= Deaths, colour= region)) +
  geom_line() +
 labs(title = "Exponential curve for the deaths in the region of Europe & Central Asia", subtitle = "Worldwide representation group by region", x= "Date",y= "Deaths", caption = "data from jh_covid19_data_2020-04-06.csv") +
   scale_x_date(date_breaks = "1 week", date_labels = "%b %d" ) +
   theme(axis.text.x = element_text(angle = 90, hjust = 1))
````

```{r fig-5, fig.cap="Figure 5", out.width="80%", fig.asp=.75,fig.align='center'}

  covid19_data %>% group_by(date,region) %>% summarize(Mean_ratio = mean(ratio)) %>%
    ggplot(aes(x=date, y= Mean_ratio, colour= region)) +
  geom_line()+
  labs(title = "We observe an increase pattern in term of the ratio  deaths/confirmed ", subtitle = "Worldwide representation groupy by region", x= "Date",y= "Deaths/Confirmed", caption = "data from jh_covid19_data_2020-04-06.csv") +
   scale_x_date(date_breaks = "1 week", date_labels = "%b %d" ) +
   theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

```{r fig-6, fig.cap="Figure 6", out.width="80%", fig.asp=.75,fig.align='center'}

covid19_data <- covid19_data %>% mutate(ratio_pop=(confirmed/population))
covid19_data$ratio[is.nan(covid19_data$ratio)]<-0

covid19_data %>% group_by(date,region) %>% summarize(Mean_ratio_pop = mean(ratio_pop)) %>%
    ggplot(aes(x=date, y= Mean_ratio_pop, colour= region)) +
  geom_line()+
  labs(title = "Europe & Central Asia \nhas the higest confirmed case by the population", subtitle = "Worldwide representation groupy by region", x= "Date",y= "Confirmed /Population", caption = "data from jh_covid19_data_2020-04-06.csv") +
   scale_x_date(date_breaks = "1 week", date_labels = "%b %d" ) +
   theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
   scale_y_continuous(labels = scales::percent)


```
 
 

Produce a worlwide map of the __COVID-19__ spread at the latest date available in `covid19_data` for each country, and describe what you see.

Hint: `ggplot2` package includes two helpful command for this part, namely

  * `map_data()` to retrieve a map,
  * and `geom_map()` to draw a map on a plot.
  
Use `expand_limits` to make sure you display the whole map of the world.







## Data selection and alignment


To compare the speed of the infection spread between countries, we need to "align" the data. In other words, we model the epidemic using equivalent "starting conditions" for every country. To do that, we filter the data so that the number of confirmed cases (in any country) is greater or equal to the maximal number of confirmed cases at the first day of a specific country. 

Let $C_{d, i}$ be the number of accumulated confirmed cases on day $d$ for country $i$. Let country $k$ be the one such that it had the highest number of reported cases on the first day in the dataset, that is $C_{1, k} \geq C_{1, i}$ for any other $i$. Find $k$ and discuss.

From `covid19_data`, extract a new table `covid19_data_filtered`:

* Select only the countries $i$, which on some day $d_{0,i}$ have $C_{d_{0,i}, i} \geq C_{0, k}$. We will call this time $d_{0,i}$ a day-zero for country $i$. In the next sections, we model $C_i(t) = C_{d_{0,i} + t,i}$, that is the spread of the epidemic in country $i$ with $t$ representing "event days". Remember, for every country the day-zero is, in general, different. However, when a country entered the epidemic stage, we are only interested in number of days that has passed from the date of entry (i.e. day-zero).
* Remove countries who are left with less than two weeks of data, i.e. we keep those countries whose number of days that has passed from the day-zero of this country is 14 or more. 
* Create a new column called `t` representing the number of days from the day-zero of this country.
* Create a new column called `confirmed_1e5pop` representing the number of confirmed cases per 100,000 habitants. This is useful in order to compare how the spread of the epidemic differs between countries relative to their population.

Discuss using either tables or visualizations or both: 

* Which countries are left in `covid19_data_filtered`?
* What is the state of the spread there?
* What is the relationship between `t` and `confirmed_1e5pop`?

```{r}
covid19_data %>%
  group_by(country) %>%
  filter(date == "2020-01-22") %>%
  arrange(desc(confirmed)) %>%
  head() %>% select(country,date, confirmed, deaths) %>%
     kable(
    caption = "List of the first countries touched by the pandemia."
    ) %>%
  kable_styling(
    bootstrap_options = "striped")



" The $k$ country is China, with 548 confirmed cases ar the earliest data in the data set. In order to do comparisons, we should create a subset where the countries' day 1 starts then the number of confirmed cases reached 548."


```

```{r}


# Filtering the data to obtain a subset containing only the first cases reported in the countries
covid19_data_filtered <- covid19_data %>%
  filter(confirmed >= 548 ) %>%
  group_by(country) %>%
     filter(min(date) %--% max(date) >=int_length(dweeks(2)))%>%
  mutate(t=as.numeric(min(date)%--%date)%/%86400)%>%
  mutate(confirmed_1e5pop= confirmed*100000/population)

   
head(covid19_data_filtered) %>% kable()

# plot of the evolution of the number of case
covid19_data_filtered%>%  filter(confirmed>=50000)%>%ggplot(aes(x=date,y=confirmed, fill=country, col=country))+geom_line()+ labs (
         y= "Confrimed case",
         title = "USA has the highest evolution of the COVID-19",
         subtitle="The evolution of confirmed case in the 7 more impacted countries" ) +
    scale_x_date(date_breaks = "1 week", date_labels = "%b %d" ) +
   theme(axis.text.x = element_text(angle = 90, hjust = 1))


# same plot but in proportion to the number of habitant 
covid19_data_filtered%>%  filter(confirmed>=50000)%>%ggplot(aes(x=date,y=confirmed_1e5pop, fill=country, col=country))+geom_line()+ labs (
         y= "confrimed case per 100'000 habitants",
         title = "Spain has the highest evolution per 100'000 habitants",
         subtitle="The evolution of confirmed case in the 7 more impacted countries" )+
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d" ) +
   theme(axis.text.x = element_text(angle = 90, hjust = 1))


" Many countries do not appear on the subset covid_data_filtered because of the threshold of 548 confirmed cases. The measures taken to prevent the covid-19 worldwide have been shared and are common among countries. This preventive measures would have been braked the spread of the virus, making more difficult to reach our threshold of 548. Furthermore, the difference in mean and in protocoles to confirm a case of coronavirus greatly varies among countries in the world. This second barrier is perhaps the greatest to overcome because of the global shortage of testing material. The reported confirmed cases might hide an unknown number of non-tested but positive case to coronavirus (Omitten/hidden cases are like false negative).

The countries removed of our panel due to our filtering are 39: Algeria, Argentina, Armenia, Azerbaijan, Bahrain, Belarus, Bosnia and Herzegovina, Cameroon, Colombia, Croatia, Dominican Republic, Egypt, Estonia, Hungary, India, Indonesia, Iraq, Kazakhstan, Kuwait, Lithuania, Mexico, Moldova, Morocco, New Zealand, North Macedonia, Panama, Peru, Philippines, Qatar, Romania, Russia, Saudi Arabia, Serbia, Singapore, Slovenia, South Africa, Tunisia, Ukraine, United Arab Emirates.

The countries included in the filtered data are 28 : Australia, Austria, Belgium, Brazil, Canada, China, Czechia, Denmark, France, Germany, Iran, Ireland, Israel, Italy, Japan, South Korea, Luxembourg, Malaysia, Netherlands, Norway, Pakistan, Portugal, Spain, Sweden, Switzerland, Turkey, United Kingdom, USA
"
```

<!--chapter:end:eda.Rmd-->

# Introduction

## The data

```{r data}
library(tidyverse)
# all the files from the data folder
files <- "data" %>%
  here::here() %>%
  list.files(full.names = TRUE)

# latest data file
latest_data <- files %>%
  str_subset("jh_covid19_data") %>%
  str_extract("\\d{4}-\\d{2}-\\d{2}") %>%
  max(na.rm = TRUE)

# load the latest data
covid19_data <- files %>%
  str_subset("jh_covid19_data") %>%
  str_subset(latest_data) %>%
  read_csv()

# load the worldbank data
worldbank_data <- files %>%
  str_subset("jh_add_wbank_data") %>%
  read_csv()

# join the covid19 and worldbank datasets
covid19_data <- covid19_data %>%
  inner_join(
    worldbank_data,
    by = "iso3c"
  )
```

For this assignment, we use COVID-19 data provided by [Johns Hopkins University (updated daily)](https://github.com/CSSEGISandData/COVID-19), as well as 
data from the world bank with demographic information.
More specifically, we use daily records of total confirmed infection cases (or cumulative number of cases), total number of fatalities (or cumulative number of deaths) per country starting from `r min(covid19_data$date)` until `r max(covid19_data$date)`. 
The dataset, downloaded on `r latest_data`, contains `r format(nrow(covid19_data), big.mark = ",")` observations and `r ncol(covid19_data)` variables (`r names(covid19_data)`).
<!-- Further, we have access to world bank data with `r nrow(worldbank_data)` rows and `r ncol(worldbank_data)` columns (`r names(worldbank_data)`) with demographic information per country. -->

## A note on Epidemiological Models

Today's epidemiological models are mostly described by so called __*SIR*__-like models [see details in @Martcheva2015, pp.9--12]. In this class of models, the population is divided into three groups:

* *(__S__)usceptible* --- people, might get infected;
* *(__I__)nfectious* --- people, who carry the infection and can infect others;
* *(__R__)ecovered/(__R__)emoved* --- people, who have already recovered from the disease and got immunity.

The SIR model is a system of ordinary nonlinear differential equations. In this homework, we focus on the following *logistic* model [see @Batista2020, pp. 2; @Martcheva2015, pp. 35--36]:

\[
  \frac{dC(t)}{dt} = r \, C(t) \cdot \left[1 - \frac{C(t)}{K}\right],
\]

where $C(t)$ is the accumulated number of cases at time $t$, $r$ is the growth rate (or infection rate), and $K$ is the final size of epidemic.
Let $C_0$ be the initial number of cases: in other words, at time $t = 0$, assume that there was $C_0$ accumulated number of cases.
The solution of the *logistic* model is

\[
  C(t) = \frac{K\cdot C_0}{C_0 + (K-C_0) \, \exp(-r\,t)},
\]

which looks like a scaled *logit* model in econometrics.

## This assigment

Because we only have access to the confirmed cases that are reported, we use those figures as a proxy for the total number of cases, with the understanding that they almost surely underestimates the actual number of interest.
In what follows, we do a preliminary exploration of the data.
We then use the *logistic* model to analyze the spread of __COVID-19__ and try to predict the final number of accumulated confirmed cases for every country. 
More specifically, we

* start by focusing on modelling the spread in Switzerland;
* then apply the same approach to every country in the dataset.


<!--chapter:end:introduction.Rmd-->

# Modeling the spread of COVID-19 worldwide

In this section, we fit the logistic model to every country in the `covid19_data_filtered` dataset.

## Fitting the logistic model to every country

Use nested data, list-columns and `logistic_model` to fit the logistic model to every country in the dataset.
Because, for some countries, the optimization method might not converge, you can use `possibly()` or something similar.
For which country does the optimization fail ?
Assess the goodness-of-fit of the logistic model in the various countries.

```{r}

by_country <- covid19_data_filtered %>% group_by(country) %>% nest()

tmp <- by_country %>%
  mutate(log_model = purrr::map(data, possibly(logistic_model, NA))) %>% 
  filter(!is.na(log_model))


fitted_model <- tmp %>% mutate(aug = purrr::map(log_model, augment)) %>% unnest(aug)
tmp %>% mutate(tdy = purrr::map(log_model, tidy)) %>% unnest(tdy)
# models <- map(by_country$data, possibly(logistic_model, NA))
# models

fitted_model %>% ggplot(aes(t,.resid)) +
  geom_line(aes(colour=country), alpha=1/3)
  
## In the model, we don't  find the Japan


```

## Fitted parameters and long-term predictions

Describe the fitted parameters (i.e., the final size and the infection rates), both on a per-country basis and some aggregate numbers (e.g., total size of the epidemic over all considered countries).
Furthermore, study the evolution (say for $t$ from 0 to 50) of the predictions of the number of confirmed cases from your models.
Similarly as was discussed in the last sub-section of the exploratory data analysis, the number of confirmed cases per 100,000 habitants is also important to understand how specific countries are managing the spread of the epidemic.
Thus, predict the evolution of this number (i.e., by dividing your predictions for confirmed cases by the population size) and discuss.

Hints:

* Format the fitted parameters using `broom::tidy()`.
* For the long-term predictions, you can use `data = data.frame(t = 0:50)` in `add_predictions()`.

```{r}
## your code goes here
```

<!--chapter:end:many_models.Rmd-->

`r if (knitr:::is_html_output()) '
# References {-}
'`

<!--chapter:end:references.Rmd-->

# Modeling the spread of COVID-19 in a single country

## The logistic model in R

Using the filtered dataset, we study the spread of COVID-19 with the logistic model.
Letting $C_{i}(t) = C_{d_{0,i} + t,i}$, the model for country $i$ can be expressed as:

\[
  C_{i}(t) = \frac{K_{i} \cdot C_{i}(0)}{C_{i}(0) + (K_{i}-C_{i}(0)) \, \exp(-R_{i}\,t)}
\]

The goal is to find the final number of cases $K_{i}$ and the infection rate $R_{i}$. We implement this in R using the following function:

```{r, echo = TRUE}
# here, we assume that data is a data frame with two variables
#  - t: the number of event days
#  - confirmed: the number of confirmed cases at time t
logistic_model <- function(data) {
  data <- data %>% arrange(t)
  C_0 <- data$confirmed[1]
  C_max <- data$confirmed[nrow(data)]
  nls(
    formula = confirmed ~ K / (1 + ((K - C_0) / C_0) * exp(-R * t)),
    data = data,
    start = list(K = 2 * C_max, R = 0.5),
    control = nls.control(minFactor = 1e-6, maxiter = 100)
  )
}
```

Notice:

* We use the nonlinear least square method `stats::nls` to fit the unknown parameters.
* In R, the formula above is `confirmed ~ K / (1 + ((K - C_0)/C_0) * exp(-R * t))`.
* As starting point $(K_0, R_0)$ for the optimiser, we set $R_0 = 0.5$ and $K_0 = 2 \, C(t^*)$, where $t^*$ is the latest information about accumulated confirmed cases.
* We further set the `control` argument as `nls.control(minFactor = 1e-6, maxiter = 100)`.

## The logistic model applied to data from Switzerland

From `covid19_data_filtered`, extract a table `covid19_ch` which corresponds to data for Switzerland.
Then:

* Use the function above to fit the logistic model for Switzerland
* Describe its output (fitted parameters, `broom::tidy()` might be useful here)
* Discuss the goodness-of-fit.
* Plot the fitted curve, as well as observed data points. 
* Present the predictions of the model. What is the estimated final size of the epidemic and infection rate in Switzerland?

```{r}
covid19_ch <- covid19_data_filtered %>% filter(country=="Switzerland")
broom::tidy(logistic_model(covid19_ch))
augment_model <-broom::augment(logistic_model(covid19_ch))
broom::augment(logistic_model(covid19_ch)) %>% print(n=10)


augment_model %>% ggplot() +
  geom_line(aes(x=t, y= confirmed, col="Observed")) +
  geom_line(aes(x=t, y= .fitted, col="Fitted" )) +
  labs(title = "Well goodness-of fit", subtitle = "Between the observed and fitted value", x= "Time period",y= "Confirmed people", col= "Model")

##Present the prediction

plot(logistic_model(covid19_ch))


```

<!--chapter:end:swiss_model.Rmd-->

