
# Exploratory data analysis

## Description of the state of the spread

Provide a high level description of the state of the spread. In particular, include

  * the number of days that have passed since the first confirmed case/death,
  * the current stage for confirmed cases/deaths/mortality (i.e., ratio of deaths to confirmed cases).
  
You can:

  * use either words, or tables, or figures, or a bit of each,
  * group your analysis per country, or region, or a bit of each.
  
  
```{r, warning=FALSE, message=FALSE, echo = FALSE}
library(tidyverse)
# all the files from the data folder
files <- "data" %>%
  here::here() %>%
  list.files(full.names = TRUE)

# latest data file
latest_data <- files %>%
  str_subset("jh_covid19_data") %>%
  str_extract("\\d{4}-\\d{2}-\\d{2}") %>%
  max(na.rm = TRUE)

# load the latest data
covid19_data <- files %>%
  str_subset("jh_covid19_data") %>%
  str_subset(latest_data) %>%
  read_csv()

# load the worldbank data
worldbank_data <- files %>%
  str_subset("jh_add_wbank_data") %>%
  read_csv()

# join the covid19 and worldbank datasets
covid19_data <- covid19_data %>%
  inner_join(
    worldbank_data,
    by = "iso3c"
  )
```

```{r, warning=FALSE, message=FALSE, echo = FALSE}
##load the library
library(tidyverse)
library(kableExtra)
library(readr)
library(stats)
library(lubridate)
library(broom)
```


Worldwide analyze

```{r fig-1, fig.cap="Figure 1", out.width="80%", fig.asp=.75,fig.align='center'}
covid19_data %>% 
  group_by(date) %>% 
  summarize(Confirmed = sum(confirmed), Deaths = sum(deaths)) %>% 
  ggplot() +
  geom_line(aes(
    x = date, 
    y = Confirmed,
    col = "Confirmed")) +
geom_line(aes(
    x = date, 
    y = Deaths, 
    col = "Deaths")) +
scale_x_date(
    date_breaks = "1 week", 
    date_labels = "%b %d" ) +
theme(
    axis.text.x = element_text(angle = 90, 
                               hjust = 1)) +
labs(
    title = "Higly increase evolution for the confirmed case",
    subtitle = "Worldwide representation",
    x = "Date",
    y = "People",
    caption = "data from jh_covid19_data_2020-04-06.csv",
    col = "Variables")
```


```{r fig-2, fig.cap="Figure 2", out.width="80%", fig.asp=.75,fig.align='center'}

covid19_data <- covid19_data %>% 
  mutate(ratio = (deaths/confirmed))
covid19_data$ratio[is.nan(covid19_data$ratio)]<-0

covid19_data %>% 
  group_by(date) %>%
  summarize(Mean_Ratio = mean(ratio)) %>%
  ggplot() +
  geom_line(aes(
    x = date, 
    y = Mean_Ratio)) +
  scale_x_date(
    date_breaks = "1 week", 
    date_labels = "%b %d" ) +
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1)) +
   labs(
     title = "High increase in the death rate related to confirmed cases of covid-19", 
     subtitle = "Worldwide representation", 
     x = "Date",
     y = "Ratio deaths/confirmed cases", 
     caption = "data from jh_covid19_data_2020-04-06.csv")
```

```{r fig-3, fig.cap="Figure 3", out.width="80%", fig.asp=.75,fig.align='center'}
 covid19_data %>%
  group_by(date,region) %>%
  summarize(Confirmed = sum(confirmed)) %>%
  ggplot(aes(
    x = date, 
    y = Confirmed, 
    colour = region)) +
  geom_line() +
  scale_x_date(
    date_breaks = "1 week", 
    date_labels = "%b %d" ) +
   theme(axis.text.x = element_text(angle = 90, 
                                    hjust = 1)) +
  labs(
    title = "Flat increase and early deperature in confirmed cases in East Asia & Pacific regions",
    subtitle = "Worldwide representation grouped by region",
    x = "Date",
    y = "Number of confirmed cases",
    caption = "data from jh_covid19_data_2020-04-06.csv",
    col = "Region") 
```

```{r fig-4, fig.cap="Figure 4", out.width="80%", fig.asp=.75,fig.align='center'}

 covid19_data %>%
  group_by(date, region) %>%
  summarize(Deaths = sum(deaths)) %>%
  ggplot(aes(
    x = date, 
    y = Deaths, 
    colour = region)) +
  geom_line() +
   scale_x_date(
     date_breaks = "1 week", 
     date_labels = "%b %d" ) +
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1)) +
  labs(
   title = "Exponential curve for the deaths in Europe & Central Asia",
   subtitle = "Worldwide representation grouped by region",
   x = "Date",
   y = "Deaths",
   caption = "data from jh_covid19_data_2020-04-06.csv",
   col = "Region") 
```

```{r fig-5, fig.cap="Figure 5", out.width="80%", fig.asp=.75,fig.align='center'}

  covid19_data %>%
  group_by(date, region) %>%
  summarize(Mean_ratio = mean(ratio)) %>%
  ggplot(aes(
    x = date, 
    y = Mean_ratio, 
    colour = region)) +
  geom_line() +
  scale_x_date(
    date_breaks = "1 week", 
    date_labels = "%b %d" ) +
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1)) +
  labs(
    title = "We observe an increase in the death rate related to confirmed cases \nof covid-19",
    subtitle = "Worldwide representation grouped by region",
    x= "Date",
    y= "Ratio deaths/confirmed cases",
    caption = "data from jh_covid19_data_2020-04-06.csv",
    col = "Region")
  
```

```{r fig-6, fig.cap="Figure 6", out.width="80%", fig.asp=.75,fig.align='center'}

covid19_data <- covid19_data %>% mutate(ratio_pop=(confirmed/population))
covid19_data$ratio[is.nan(covid19_data$ratio)]<-0

covid19_data %>%
  group_by(date,region) %>%
  summarize(Mean_ratio_pop = mean(ratio_pop)) %>%
  ggplot(aes(
    x = date, 
    y = Mean_ratio_pop, 
    colour = region)) +
  geom_line() +
  scale_x_date(
    date_breaks = "1 week", 
    date_labels = "%b %d" ) +
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1)) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Europe & Central Asia are the regions with the highest \ncontamination rate related to covid-19...",
    subtitle = "Worldwide representation grouped by region",
    x = "Date",
    y = "Ratio confirmed cases / population size",
   caption = "data from jh_covid19_data_2020-04-06.csv",
   col = "Region")
  
```

```{r fig-7, fig.cap="Figure 7", out.width="80%", fig.asp=.75,fig.align='center'}

covid19_data <- covid19_data %>% 
  mutate(ratio_death_pop = (deaths/population))
covid19_data$ratio[is.nan(covid19_data$ratio_death_pop)]<-0

covid19_data %>%
  group_by(date, region) %>%
  summarize(Mean_ratio_death_pop = mean(ratio_death_pop)) %>%
  ggplot(aes(
    x = date, 
    y = Mean_ratio_death_pop, 
    colour = region)) +
  geom_line() +
  scale_x_date(
    date_breaks = "1 week", 
    date_labels = "%b %d" ) +
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1)) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "... Europe & Central Asia face as well the highest \ndeath ratio related to covid-19",
    subtitle = "Worldwide representation grouped by region",
    x = "Date",
    y = "Ratio deaths / population size",
   caption = "data from jh_covid19_data_2020-04-06.csv",
   col = "Region") 
  
```
 

Produce a worlwide map of the __COVID-19__ spread at the latest date available in `covid19_data` for each country, and describe what you see.

Hint: `ggplot2` package includes two helpful command for this part, namely

  * `map_data()` to retrieve a map,
  * and `geom_map()` to draw a map on a plot.
  
Use `expand_limits` to make sure you display the whole map of the world.


```{r fig-8, fig.cap="Figure 8", out.width="80%", fig.asp=.75,fig.align='center'}

covid <- read_csv(file = here::here("data/jh_covid19_data_2020-04-06.csv"))

#filter by the latest data available for each country
covid_latest =covid %>% group_by(country)%>% filter(date == max(date)
  )

#add long lat to table
worldmap <- map_data("world")
worldmap= worldmap %>% 
  rename(country=region)

graph1= left_join(worldmap, covid_latest,"country")

#plot the table
  ggplot(graph1, aes(long, lat,group=group))+
  geom_polygon(aes(fill = confirmed ), color = "white")+
  scale_fill_viridis_c(option = "C", labels = scales::comma)+labs(title = "Number of people Confirmed COVID-19 at the latest date available")+ theme(legend.text = element_text(colour="blue", size=6, face="bold"))+ expand_limits(x =graph1$long, y = graph1$lat)

```

## Data selection and alignment


To compare the speed of the infection spread between countries, we need to "align" the data. In other words, we model the epidemic using equivalent "starting conditions" for every country. To do that, we filter the data so that the number of confirmed cases (in any country) is greater or equal to the maximal number of confirmed cases at the first day of a specific country. 

Let $C_{d, i}$ be the number of accumulated confirmed cases on day $d$ for country $i$. Let country $k$ be the one such that it had the highest number of reported cases on the first day in the dataset, that is $C_{1, k} \geq C_{1, i}$ for any other $i$. Find $k$ and discuss.

From `covid19_data`, extract a new table `covid19_data_filtered`:

* Select only the countries $i$, which on some day $d_{0,i}$ have $C_{d_{0,i}, i} \geq C_{0, k}$. We will call this time $d_{0,i}$ a day-zero for country $i$. In the next sections, we model $C_i(t) = C_{d_{0,i} + t,i}$, that is the spread of the epidemic in country $i$ with $t$ representing "event days". Remember, for every country the day-zero is, in general, different. However, when a country entered the epidemic stage, we are only interested in number of days that has passed from the date of entry (i.e. day-zero).
* Remove countries who are left with less than two weeks of data, i.e. we keep those countries whose number of days that has passed from the day-zero of this country is 14 or more. 
* Create a new column called `t` representing the number of days from the day-zero of this country.
* Create a new column called `confirmed_1e5pop` representing the number of confirmed cases per 100,000 habitants. This is useful in order to compare how the spread of the epidemic differs between countries relative to their population.

Discuss using either tables or visualizations or both: 

* Which countries are left in `covid19_data_filtered`?
* What is the state of the spread there?
* What is the relationship between `t` and `confirmed_1e5pop`?

```{r}
covid19_data %>%
  group_by(country) %>%
  filter(date == "2020-01-22") %>%
  arrange(desc(confirmed)) %>%
  head() %>%
  select(country, date, confirmed, deaths) %>%
     kable(
    caption = "List of the first countries touched by the pandemic."
       ) %>%
  kable_styling(
    bootstrap_options = "striped")


" The $k$ country is China, with 548 confirmed cases ar the earliest data in the data set. In order to do comparisons, we should create a subset where the countries' day 1 starts when the number of confirmed cases reached 548."

```

```{r}

# Filtering the data to obtain a subset containing only the first cases reported in the countries
covid19_data_filtered <- covid19_data %>%
  filter(confirmed >= 548 ) %>%
  group_by(country) %>%
     filter(min(date) %--% max(date) >= int_length(dweeks(2)))%>%
  mutate(t = as.numeric(min(date)%--%date)%/%86400)%>%
  mutate(confirmed_1e5pop = confirmed*100000/population)
```

```{r}   
head(covid19_data_filtered) %>%
  select(country, date, confirmed,, deaths, population) %>%
     kable(
    caption = "Glimpse of the countries' day 1 aligned"
       ) %>%
  kable_styling(
    bootstrap_options = "striped")
```

```{r fig-9, fig.cap="Figure 9", out.width="80%", fig.asp=.75,fig.align='center'}
# plot of the evolution of the number of case
covid19_data_filtered %>% 
  filter(confirmed >= 50000) %>%
  ggplot(aes(
    x = date,
    y = confirmed, 
    fill = country, 
    col = country)) +
  geom_line() +
  scale_x_date(
    date_breaks = "1 week", 
    date_labels = "%b %d" ) +
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1)) +
  labs(
    x = "Date",
    y = "Confirmed cased",
    title = "USA has the highest evolution of the COVID-19",
    subtitle = "The evolution of confirmed cases in the 7 most impacted countries")
```

```{r fig-10, fig.cap="Figure 10", out.width="80%", fig.asp=.75,fig.align='center'}
# same plot but in proportion to the number of habitant 
covid19_data_filtered %>% 
  filter(confirmed>=50000) %>%
  ggplot(aes(
    x = date,
    y = confirmed_1e5pop, 
    fill = country, 
    col = country)) +
  geom_line() +
  scale_x_date(
    date_breaks = "1 week",
    date_labels = "%b %d") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  labs (
    x = "Date",
    y = "confirmed cases per 100'000 inhabitants",
    title = "Spain has the highest evolution per 100'000 habitants",
    subtitle = "The evolution of confirmed cases in the 7 most impacted countries" )
```

" Many countries do not appear on the subset covid_data_filtered because of the threshold of 548 confirmed cases. The measures taken to prevent the covid-19 worldwide have been shared and are common among countries. This preventive measures would have been braked the spread of the virus, making more difficult to reach our threshold of 548. Furthermore, the difference in mean and in protocoles to confirm a case of coronavirus greatly varies among countries in the world. This second barrier is perhaps the greatest to overcome because of the global shortage of testing material. The reported confirmed cases might hide an unknown number of non-tested but positive case to coronavirus (Omitten/hidden cases are like false negative).

The countries removed of our panel due to our filtering are 39: Algeria, Argentina, Armenia, Azerbaijan, Bahrain, Belarus, Bosnia and Herzegovina, Cameroon, Colombia, Croatia, Dominican Republic, Egypt, Estonia, Hungary, India, Indonesia, Iraq, Kazakhstan, Kuwait, Lithuania, Mexico, Moldova, Morocco, New Zealand, North Macedonia, Panama, Peru, Philippines, Qatar, Romania, Russia, Saudi Arabia, Serbia, Singapore, Slovenia, South Africa, Tunisia, Ukraine, United Arab Emirates.

The countries included in the filtered data are 28 : Australia, Austria, Belgium, Brazil, Canada, China, Czechia, Denmark, France, Germany, Iran, Ireland, Israel, Italy, Japan, South Korea, Luxembourg, Malaysia, Netherlands, Norway, Pakistan, Portugal, Spain, Sweden, Switzerland, Turkey, United Kingdom, USA
"
