# Modeling the spread of COVID-19 worldwide

In this section, we fit the logistic model to every country in the `covid19_data_filtered` dataset.

## Fitting the logistic model to every country

Use nested data, list-columns and `logistic_model` to fit the logistic model to every country in the dataset.
Because, for some countries, the optimization method might not converge, you can use `possibly()` or something similar.
For which country does the optimization fail ?
Assess the goodness-of-fit of the logistic model in the various countries.

```{r}
## your code goes here
```

## Fitted parameters and long-term predictions

Describe the fitted parameters (i.e., the final size and the infection rates), both on a per-country basis and some aggregate numbers (e.g., total size of the epidemic over all considered countries).
Furthermore, study the evolution (say for $t$ from 0 to 50) of the predictions of the number of confirmed cases from your models.
Similarly as was discussed in the last sub-section of the exploratory data analysis, the number of confirmed cases per 100,000 habitants is also important to understand how specific countries are managing the spread of the epidemic.
Thus, predict the evolution of this number (i.e., by dividing your predictions for confirmed cases by the population size) and discuss.

Hints:

* Format the fitted parameters using `broom::tidy()`.
* For the long-term predictions, you can use `data = data.frame(t = 0:50)` in `add_predictions()`.

```{r}
## your code goes here
```
