# Modeling the spread of COVID-19 in a single country

## The logistic model in R

Using the filtered dataset, we study the spread of COVID-19 with the logistic model.
Letting $C_{i}(t) = C_{d_{0,i} + t,i}$, the model for country $i$ can be expressed as:

\[
  C_{i}(t) = \frac{K_{i} \cdot C_{i}(0)}{C_{i}(0) + (K_{i}-C_{i}(0)) \, \exp(-R_{i}\,t)}
\]

The goal is to find the final number of cases $K_{i}$ and the infection rate $R_{i}$. We implement this in R using the following function:

```{r, echo = TRUE}
# here, we assume that data is a data frame with two variables
#  - t: the number of event days
#  - confirmed: the number of confirmed cases at time t
logistic_model <- function(data) {
  data <- data %>% arrange(t)
  C_0 <- data$confirmed[1]
  C_max <- data$confirmed[nrow(data)]
  nls(
    formula = confirmed ~ K / (1 + ((K - C_0) / C_0) * exp(-R * t)),
    data = data,
    start = list(K = 2 * C_max, R = 0.5),
    control = nls.control(minFactor = 1e-6, maxiter = 100)
  )
}
```

Notice:

* We use the nonlinear least square method `stats::nls` to fit the unknown parameters.
* In R, the formula above is `confirmed ~ K / (1 + ((K - C_0)/C_0) * exp(-R * t))`.
* As starting point $(K_0, R_0)$ for the optimiser, we set $R_0 = 0.5$ and $K_0 = 2 \, C(t^*)$, where $t^*$ is the latest information about accumulated confirmed cases.
* We further set the `control` argument as `nls.control(minFactor = 1e-6, maxiter = 100)`.

## The logistic model applied to data from Switzerland

From `covid19_data_filtered`, extract a table `covid19_ch` which corresponds to data for Switzerland.
Then:

* Use the function above to fit the logistic model for Switzerland
* Describe its output (fitted parameters, `broom::tidy()` might be useful here)
* Discuss the goodness-of-fit.
* Plot the fitted curve, as well as observed data points. 
* Present the predictions of the model. What is the estimated final size of the epidemic and infection rate in Switzerland?


The prediction shows that the final size of the confirmed case is about 21'500 cases, with an infection rate of 0.246%.
```{r table 3, warning=FALSE, message=FALSE, echo = FALSE}
covid19_ch <- covid19_data_filtered %>% filter(country=="Switzerland")
broom::tidy(logistic_model(covid19_ch))  %>%
     kable(
    caption = "K = confirmed case, R = Infesction rate"
       ) %>%
  kable_styling(
    bootstrap_options = "striped")
```


```{r table 4, warning=FALSE, message=FALSE, echo = FALSE}
augment_model <-broom::augment(logistic_model(covid19_ch))
broom::augment(logistic_model(covid19_ch)) %>% print(n=10) %>%
     kable(
    caption = "Presentation of the prediction"
       ) %>%
  kable_styling(
    bootstrap_options = "striped")
```


In this following plot, we observe  the fitted curve, as well as observed data points
```{r fig-12, fig.cap="Figure 12", out.width="80%", fig.asp=.75,fig.align='center', warning=FALSE, message=FALSE, echo = FALSE}
augment_model %>% ggplot() +
  geom_line(aes(x=t, y= confirmed, col="Observed")) +
  geom_line(aes(x=t, y= .fitted, col="Fitted" )) +
  labs(title = "Well goodness-of fit", subtitle = "Between the observed and fitted value", x= "Time period",y= "Confirmed people", col= "Model")
```
The 2 curves are close, which mean that our predictions are well estimate, regarding are true value.


In order to well assess our model, we pot the residuals of the mode.

```{r fig-13, fig.cap="Figure 13", out.width="80%", fig.asp=.75,fig.align='center', warning=FALSE, message=FALSE, echo = FALSE}
logistic_model(covid19_ch)

ggplot(augment_model) + 
  geom_point(aes(x=.fitted, y=.resid)) +
  labs(
    title = "Residuals increase regarding the increase of our prediction",
    x= "Prediction",
    y= "Residuals"
  )
```
We observe a pretty well prediction when our fitted value are small, but when they increase, our residuals increase too.