# Modeling the spread of COVID-19 in a single country

## The logistic model in R

Using the filtered dataset, we study the spread of COVID-19 with the logistic model.
Letting $C_{i}(t) = C_{d_{0,i} + t,i}$, the model for country $i$ can be expressed as:

\[
  C_{i}(t) = \frac{K_{i} \cdot C_{i}(0)}{C_{i}(0) + (K_{i}-C_{i}(0)) \, \exp(-R_{i}\,t)}
\]

The goal is to find the final number of cases $K_{i}$ and the infection rate $R_{i}$. We implement this in R using the following function:

```{r, echo = TRUE}
# here, we assume that data is a data frame with two variables
#  - t: the number of event days
#  - confirmed: the number of confirmed cases at time t
logistic_model <- function(data) {
  data <- data %>% arrange(t)
  C_0 <- data$confirmed[1]
  C_max <- data$confirmed[nrow(data)]
  nls(
    formula = confirmed ~ K / (1 + ((K - C_0) / C_0) * exp(-R * t)),
    data = data,
    start = list(K = 2 * C_max, R = 0.5),
    control = nls.control(minFactor = 1e-6, maxiter = 100)
  )
}
```

Notice:

* We use the nonlinear least square method `stats::nls` to fit the unknown parameters.
* In R, the formula above is `confirmed ~ K / (1 + ((K - C_0)/C_0) * exp(-R * t))`.
* As starting point $(K_0, R_0)$ for the optimiser, we set $R_0 = 0.5$ and $K_0 = 2 \, C(t^*)$, where $t^*$ is the latest information about accumulated confirmed cases.
* We further set the `control` argument as `nls.control(minFactor = 1e-6, maxiter = 100)`.

## The logistic model applied to data from Switzerland

From `covid19_data_filtered`, extract a table `covid19_ch` which corresponds to data for Switzerland.
Then:

* Use the function above to fit the logistic model for Switzerland
* Describe its output (fitted parameters, `broom::tidy()` might be useful here)
* Discuss the goodness-of-fit.
* Plot the fitted curve, as well as observed data points. 
* Present the predictions of the model. What is the estimated final size of the epidemic and infection rate in Switzerland?

```{r}
covid19_ch <- covid19_data_filtered %>% filter(country=="Switzerland")
broom::tidy(logistic_model(covid19_ch))
augment_model <-broom::augment(logistic_model(covid19_ch))
broom::augment(logistic_model(covid19_ch)) %>% print(n=10)

augment_model <-broom::augment(logistic_model(covid19_ch))
##The prediction shows that the final size of the confirmed case is about 21'500 cases, with an infection rate of 0.246%.


augment_model %>% ggplot() +
  geom_line(aes(x=t, y= confirmed, col="Observed")) +
  geom_line(aes(x=t, y= .fitted, col="Fitted" )) +
  labs(title = "Well goodness-of fit", subtitle = "Between the observed and fitted value", x= "Time period",y= "Confirmed people", col= "Model")

##Present the prediction

plot(logistic_model(covid19_ch))


```
